\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-09 NAR_DRAFT.tex
add thing about copy number calls in the discussion

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
\bkmcomment{We make two simplifying assumptions here.
First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
Second, we assume 100\% tumor purity.
MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
We encourage users to take advantage of the MuTect contamination correction where appropriate.
In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
}
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.

@@

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.

>>>>

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
\bkmcomment{We make two simplifying assumptions here.
First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
Second, we assume 100\% tumor purity.
MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
We encourage users to take advantage of the MuTect contamination correction where appropriate.
In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
}
%We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.

@@

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
\bkmcomment{We use properties of the allele frequency distribution generated by mutation acquisition during exponential growth.
Copy number alterations can distort the allele frequency distribution.
While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}
Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
